---
title: "Forecast daily bike rental demand using time series models"
date: "`r Sys.Date()`"
output: html_document
author: "Jaishree Joshita"
---

# About Data Analysis Report

This RMarkdown file contains the report of the data analysis done for the project on forecasting daily bike rental demand using time series models in R. It contains analysis such as data exploration, summary statistics and building the time series models. The final report was completed on `r date()`. 

**Data Description:**

This dataset contains the daily count of rental bike transactions between years 2011 and 2012 in Capital bikeshare system with the corresponding weather and seasonal information.

**Data Source:** https://archive.ics.uci.edu/ml/datasets/bike+sharing+dataset

**Relevant Paper:** 

Fanaee-T, Hadi, and Gama, Joao. Event labeling combining ensemble detectors and background knowledge, Progress in Artificial Intelligence (2013): pp. 1-15, Springer Berlin Heidelberg



# Task One: Load and explore the data

## Load data and install packages

```{r}
## Import required packages
install.packages("tidyverse")
library("tidyverse")
library("ggplot2")
library("dbplyr")
install.packages("timetk")
install.packages("tseries")
install.packages("forecast")
library("timetk")
library("tseries")
library("forecast")

#setting directory path for file access
#getwd()
#setwd("~/Users/jaishreejoshita/Documents/Bike Sharing Dataset in R/bike sharing dataset")

#reading the file
day_data <- read.csv("day.csv")
summary(day_data)
```

## Describe and explore the data

```{r}
day_data[,"dteday"] <- as.Date(day_data[,"dteday"]) #This line ensures that the dteday column is interpreted as dates, which allows for date-specific operations and plotting.
day_data[,"ncnt"] <- day_data[,"cnt"] / max(day_data[,"cnt"]) #This creates a new column ncnt where each value is the original cnt value divided by the maximum cnt value in the dataset. This scales the cnt values to a range of 0 to 1.
day_data[,"nr"] <- day_data[,"registered"] / max(day_data[,"registered"]) #his creates a new column nr with the registered values normalized to a range of 0 to 1.
day_data[,"rr"] <- day_data[,"cnt"] / max(day_data[,"registered"]) #This creates a new column rr where each value is the original cnt value divided by the maximum registered value. This column represents the ratio of total counts to the maximum registered users.
summary(day_data)
```

# Task Two: Create interactive time series plots

```{r}
## Read about the timetk package
#?timetk
#The timetk package's plot_time_series function is powerful and allows for a lot of customization. You can refer to the timetk documentation for more details and customization options.
install.packages("timetk")
library("timetk")
#The plot is interactive, and the x and y axes, along with the title, are labeled appropriately. This is accomplished using functions from the dplyr and timetk packages.
```


# Task Three: ?Smooth time series data

```{r}
#day_data
install.packages("dplyr")
library(dplyr)

# Ensure dteday is in Date format
day_data <- day_data %>%
  mutate(dteday = as.Date(dteday))

# Group by 'yr' and create the time series plot
day_data %>%
  group_by(yr) %>%
  plot_time_series(
    .date_var = dteday,
    .value = temp,
    .color_var = season,
    .x_lab = "Date",
    .y_lab = "Normalized Temperature",
    .title = "Normalized Temperature vs Date for Day Data",
    .interactive = TRUE)

day_data %>%
  group_by(yr) %>%
  plot_time_series(
    .date_var = dteday,
    .value = hum,
    .color_var = season, 
    .x_lab = "Date", 
    .y_lab = "Normalized Humidity",
    .title = "Normalized Humidity vs Date for Day Data", 
    .interactive = TRUE)

day_data %>% 
  group_by(yr) %>% 
  plot_time_series(
    .date_var = dteday,
    .value = windspeed, 
    .color_var = season, 
    .x_lab = "Date", 
    .y_lab = "Normalized Windspeed", 
    .title = "Windspeed vs Date for Day Data", 
    .interactive = TRUE)

day_data %>% 
  group_by(yr) %>% 
  plot_time_series(
    .date_var = dteday,
    .value = ncnt,
    .color_var = season, 
    .x_lab = "Date", 
    .y_lab = "Normalized Bike Rentals", 
    .title = "Normalized Bike Rentals vs Date for Day Data", 
    .interactive = TRUE)

day_data %>% 
  group_by(., yr) %>% 
  plot_time_series(
   .date_var = dteday,
    .value = nr, 
    .color_var = season, 
    .x_lab = "Date", 
    .y_lab = "Normalized Registered Rentals", 
    .title = "Normalized Registered Rentals vs Date for Day Data", 
    .interactive = TRUE)

day_data %>% 
  group_by(yr) %>% 
  plot_time_series(
   .date_var = dteday,
    .value = rr, 
    .color_var = season, 
    .x_lab = "Date", 
    .y_lab = "Normalized Registered Rentals", 
    .title = "Ratio of Rentals to Registration vs Date for Day Data", 
    .interactive = TRUE)
```



# Task Four: Decompose and assess the stationarity of time series data

```{r}
# Perform ADF test on the 'temp' column
#day data
day_data[,"temp"] %>% 
adf.test()
#or
adf_test_result <- adf.test(day_data$temp)
print(adf_test_result)

day_data[,"ncnt"] %>% adf.test()
day_data[,"nr"] %>% adf.test()
day_data[,"rr"] %>% adf.test()

#decomposes the data
freq <- 365

#normalized day rentals
norm_rentals <- ts(day_data[, "nr"], frequency = freq)
decomped1 <- stl(norm_rentals, "periodic")
plot(decomped1$time.series[,2], ylab = "Stationary of the Normalized Rental Reservations", 
     xlab = "Day of the Year")

checkresiduals(decomped1$time.series[, 3])

#normalized day counts
norm_cnt <- ts(day_data[, "ncnt"], frequency = freq)
decomped2 <- stl(norm_cnt, "periodic")
plot(decomped2$time.series[,2], ylab = "Stationary of the Normalized Rental Counts", 
     xlab = "Day of the Year")

checkresiduals(decomped2$time.series[, 3])

#normalized day rental rates
norm_rr <- ts(day_data[, "rr"], frequency = freq)
decomped3 <- stl(norm_rr, "periodic")
plot(decomped3$time.series[,2], ylab = "Stationary of the Normalized Rental Counts to Reservations", 
     xlab = "Day of the Year")

checkresiduals(decomped3$time.series[, 3])

#returns the stats
print("-------Noramlized Rental Reservations")

shapiro.test(decomped1$time.series[, 3])

print("-------Normalized Count of Rentals")

shapiro.test(decomped2$time.series[, 3])

print("-------Normalized Ratio of Rentals to Reservations")
shapiro.test(decomped3$time.series[, 3])

```



# Task Five: Fit and forecast time series data using ARIMA models

```{r}
#bike count predictions
fit1 <- auto.arima(norm_cnt, seasonal = TRUE, )
hist(fit1$residuals, xlab = "Residual", ylab = "Distribution", main = "Histogram of Model Errors - Bike Count")
shapiro.test(fit1$residuals)

prediction1 <- forecast(fit1, 25)
plot(prediction1, xlab = "Date", ylab = "Normalized Count of Rentals", main = "Prediction of Bike Rental Counts")

#predictions of number of reservations
fit2 <- auto.arima(norm_rentals, seasonal = TRUE, )
hist(fit2$residuals, xlab = "Residual", ylab = "Distribution", main = "Histogram of Model Errors - Rental Count")
shapiro.test(fit2$residuals)

prediction2 <- forecast(fit2, 25)
plot(prediction2, xlab = "Date", ylab = "Normalized Rentals", main = "Prediction of Bike Rentals")

#bike count predictions
fit3 <- auto.arima(norm_cnt, seasonal = TRUE, )
hist(fit2$residuals, xlab = "Residual", ylab = "Distribution", main = "Histogram of Model Errors - Count to Rental Ratio")
shapiro.test(fit3$residuals)

prediction3 <- forecast(fit2, 25)
plot(prediction3, xlab = "Date", ylab = "Normalized Rental Ratio", main = "Prediction of Bike Rentals to Reservations")

```



# Task Six: Findings and Conclusions

After processing the raw data and using the ARIMA package to model ride share data, I was able to make predictions for the 25 days beyond the current data set. Qualitatively the data shows that was the weather gets warmer the number of bike rentals increase, and over the course of two years the number of rentals increases over the number of rentals from the previous year. As the data terminates at the end of one cycle, I expect the number of rentals to increase to a level higher than it was a year before, which is what the models are predicting. Therefore the results were what I expected - the data appears to oscillate up and down over a 1 year period with the overall data moving towards higher rental numbers.































